Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array

### 2023-07-04 docs/manuscript/paper.tex
implement AM corrections

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{State-space PTA}

@@

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
editorials, notices -- miscellaneous
\end{keywords}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Wyithe2003,Sesana2013,Ravi2015MNRAS.447.2772R},  would result in the emission of nHz gravitational waves (GWs). Other potential GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of these long-wavelength GWs has necessitated the development of novel astrophysical methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost technique for the detection of GWs in this  low-frequency regime is a via timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{2021hgwa.bookE...4V}. The presence  of a nHz GW will influence the propagation of the pulsar radio beacon, leaving a characteristic impression on the received pulsar timing signal. By measuring the modulation of the pulsar signal in this way, one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have now been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{2020ApJ...905L..34A}, the Parkes Pulsar Timing array \citep[PPTA][]{2020PASA...37...20K}, and the European Pulsar Timing Array \citep[EPTA,][]{2010CQGra..27h4014F}. These previously disparate efforts have now been joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, as MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on the detection of this stochastic background via measuring the correlation in the pulsar timing residuals between any two pair of pulsars; the presence of a GW induces a characteristic correlation as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}. SMBHBs which are sufficiently massive and nearby may be individually resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
The stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. However, as with the stochastic background, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Zhu2014PPTA,Babak2016,Arzoumanian2023}. \newline

The intrinsic pulsar timing noise - i.e.  random, unmodelled variations in the pulse arrival time due to irregularities in the rotation of the star - has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in both the internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J}, magnetospheric state switching \citep{magneto1,Lyne2010L} and superfluid turbulence \citep{Link10.111,Melatos2014}. In order to mitigate the influence of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are known to be much more stable rotators with minimal timing noise. However, timing noise in MSPs could be a ``latent" phenomenon that will emerge with the increased detector sensitivity and longer timing spans necessary to detect GWs \citep{Shannon2010}. In addition to the temporally uncorrelated ``red" timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present a novel approach to PTA data analysis for individual, monochromatic, SMBHB GW sources that can self-consistently account for the intrinsic timing noise inherent to pulsar observations. We formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution by using a Kalman filter 
\citep{Kalman1,Meyers2021,Melatos2023}.  With the tracking of the intrinsic pulsar state established, it is then possible to determine the presence of a GW signal in the pulsar data and infer the underlying source parameters. \newline

\noindent This paper is organised as follows. In Section \ref{sec:model} we present our state-space model for the pulsar pulse frequency which is subject the influence of a GW. In Section \ref{sec:detect} we discretise the model, develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence. In Section \ref{sec:testing} we test this new method on synthetic pulsar data. Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature. \newline

@@

\section{State-Space Frequency Model}\label{sec:model}
We want to formulate the PTA analysis as a state-space problem with a separation between the intrinsic pulsar state, and the measurement state recorded by an observer. For this work we will take our state variable to be the intrinsic pulsar pulse frequency $f_{\rm p}(t)$, as measured in the momentarily comoving reference frame of an observer local to the pulsar. We will take our measurement variable to be the pulsar pulse frequency that is measured by an observer at Earth, $f_{\rm m}(t)$.  In Sections \ref{sec:psr_frequency}, \ref{sec:psr_measured} we will describe our model for $f_{\rm p}(t)$ and  $f_{\rm m}(t)$ respectively.

We will go on in Section \ref{sec:statespace} to bring together these two models into a cohesive, discretized state space model. 
\textcolor{red}{TK: Need some extra discussion here on why this choice of state/measurement space. Why not e.g. phase, TOA, voltage, etc. ?}

\subsection{Evolution of the pulsar frequency} \label{sec:psr_frequency}
We will take as our model of the intrinsic pulsar frequency $f_{\rm p}(t)$ a variation of the phenomenological model of \cite{Vargas}. Within this model, $f_{\rm p}(t)$ evolves according to a combination of both deterministic torques (i.e. electromagnetic spin-down) and stochastic torques (i.e. `spin wandering', achromatic variations in the pulse TOA intrinsic to the star). The deterministic torque is taken to arise from the pulsar magnetic dipole, with braking index $n=3$ whilst the stochastic torque is a simple white noise process. Specifically, the frequency evolves according to a Ornstein-Uhlenbeck process (equivalently a Langevin equation) with a time-dependent drift parameter:
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p} - f_{\rm EM} (t)] + \dot{f}_{\rm EM} + \xi(t)
	\label{eq:frequency_evolution}
\end{equation}
where $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma$ a proportionality constant that specifies the mean-reversion timescale, and $\xi(t)$ a white noise process that satisfies,
\begin{equation}
	\langle \xi(t) \xi(t') \rangle = \sigma^2 \delta(t - t')
\end{equation}
for variance $\sigma^2$. For PTA analysis, we are concerned with timescales on the order of years. Consequently, we can express the EM spindown straightforwardly as
\begin{equation}
	f_{\rm EM} (t) = f_{\rm EM}(0) + \dot{f}_{\rm EM} (0)t
\end{equation}  
Completely, the frequency evolution is then given by the solution of the stochastic differential equation,
\begin{equation}
	\frac{df_{\rm p}}{dt} = -\gamma	 [f_{\rm p}- f_{\rm EM}(0) - \dot{f}_{\rm EM} (0)t] + \dot{f}_{\rm EM} (0) + \xi(t)
	\label{eq:frequency_evolution_expanded}
\end{equation}
As emphasised in \cite{Vargas}, this model for the frequency evolution is a phenomenological model that aims to qualitatively reproduce the typical behaviour of observed pulsars, rather than being derived from a physical model of the neutron star \citep[e.g. a model of the neutron star crust and superfluid, components][]{Meyers2021}. For our purposes of exploring the detection of GWs via a state space formulation it will prove sufficiently accurate and appropriate.

\subsection{Modulation of pulsar frequency due to a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local neutron star reference frame is different from that measured by an observer on Earth. We will now derive the influence of the GW on the received pulse frequency.

\subsubsection{Plane GW perturbation}
We take a gravitational plane wave that perturbs a background Minkowski spacetime as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
for Minkowski metric $ \eta_{\mu \nu}$, spatial coordinates $\bar{x}$, where the GW has a constant angular frequency $\Omega$, propagates in the $\bar{n}$-direction and has a phase offset of  $\Phi_0$. We emphasise that for this work $\Omega$ has no time dependence - we are concerned solely with monochromatic sources. We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the Earth. The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$) whilst the spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\bar{n}) + h_{\times} e_{ij}^{\times}(\bar{n})
\end{align}
where  $h_{+,\times}$ are the polarisation amplitudes of the gravitational plane wave which are given by
\begin{align}
	h_+ &= h(1 + \cos^2 \iota) \\
		h_{\times} &= -2h\cos \iota
\end{align}
The polarisation tensors $e_{ij}^{+,\times}$ are uniquely defined by the principal axes of the wave:
\begin{align}
	e_{a b}^{+}(\hat{\Omega}) =\hat{m}_a \hat{m}_b-\hat{n}_a \hat{n}_b \\
		e_{a b}^{\times}(\hat{\Omega}) =\hat{m}_a \hat{n}_b+\hat{n}_a \hat{m}_b
\end{align}
which are in turn specified via the location of the GW source on the sky (via polar coordinate $\theta$ and azimuthal coordinate $\phi$) and the polarisation angle $\psi$
\begin{align}
	\vec{m} & =(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \hat{y} \nonumber \\
	& +(\sin \psi \sin \theta) \hat{z} \\
	\vec{n} & =(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \hat{x} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \hat{y}\nonumber  \\
	& +(\cos \psi \sin \theta) \hat{z}
\end{align}

\subsubsection{Pulse frequency as a photon}
\noindent We will consider the pulse frequency as a photon with covariant 4-momentum $p_{\mu}$. Generally, the frequency of a photon recorded by an observer with 4-velocity $u^{\mu}$ is 
\begin{equation}
	f = p_{\alpha} u^{\alpha}
	\label{eq:freq_temporal}
\end{equation}
\noindent We consider both our emitter and receiver to be stationary, such that  
\begin{equation}
	u^{\alpha}|_{\rm emitter} = u^{\alpha}|_{\rm receiver} = (1,0,0,0)
\end{equation}
\noindent Consequently the frequency can be directly identified with the temporal component of the covariant 4-momentum,
\begin{equation}
	f = p_t
\end{equation}
\noindent The expression for the evolution of the pulse frequency as measured by the observer on Earth is then,
\begin{equation}
	p_t(t_1)|_{\rm Earth} = p_t(t_0)|_{\rm source} + \int_{t = t_0}^{t=t_1} \dot{p}_t dt
	\label{eq:integral}
\end{equation}
\noindent where the overdot denotes a derivative w.r.t. $t$. Since the influence of the GW perturbation on $\dot{p}_t$ is small, we can relate the source emission and receiver times as $t_1 = t_0 + d$ and consider the photon trajectory to be an unperturbed path.

\subsubsection{Hamiltonian Mechanics}
The Hamiltonian in covariant notation can be written as 
\begin{equation}
	H = \frac{1}{2} g_{\mu \nu} p^{\mu} p^{\nu},
\end{equation}
\noindent which if we substitute in our expression for the perturbed metric is
\begin{equation}
	H = \frac{1}{2} \eta_{\mu \nu} p^{\mu} p^{\nu} + \frac{1}{2} H_{ij}p^i p^j e^{i(\Omega(\bar{n} \cdot \bar{x} - t) + \Phi_0)	}
\end{equation}
\noindent Now, Hamilton's equations are
\begin{equation}
	\frac{dx^{\mu}}{d\lambda} = \frac{\partial H}{\partial p_{\mu}} , \, \, \frac{dp_{\mu}}{d \lambda} = -\frac{\partial H}{\partial x^{\mu}} 
\end{equation}

\noindent for affine parameter $\lambda$. The derivative of the temporal component of the covariant momenta is then,
\begin{equation}
	\frac{d p_{t}}{d \lambda} = -\frac{i\Omega}{2} H_{ij}p^i p^j  e^{i(\Omega(\bar{n}\cdot \bar{x} - t)+\Phi_0)}
\end{equation}
\noindent The derivative w.r.t coordinate time $t$ is
\begin{equation}
	\dot{p}_t = \frac{d p_{t}}{d \lambda} \left(\frac{dt}{d\lambda}\right)^{-1} = \frac{d p_{t}}{d \lambda} \left(\frac{1}{p^t}\right)
\end{equation}
\noindent Note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. 
It will prove useful to recognise that 
\begin{equation}
	p^{\mu} = \omega(1,-q^x,-q^y,-q^z)
\end{equation} 
where $\bar{q}$ is the unit vector between the Earth and pulsar and $\omega$ is the \textit{constant} photon angular frequency. Given the small effect of the GW perturbation, at first order we can identify $\omega$ as either the frequency at source or observer. Similarly, we can parameterize the photon spatial coordinates $\bar{x}$ as,
\begin{equation}
	\bar{x}(t) = - \bar{q} (t - t_1)
\end{equation} 
Note that when our integration variable $t = t_1$ (i.e. the photon is at Earth) then $\bar{x}(t=t_1) = 0$ which is what we expect given our chosen coordinate system. Also note that $\dot{p}_t$ is entirely a function of the GW perturbation. In the Minkowski case the spacetime is stationary and so $p_t$ should be conserved along the geodesic. We will treat the pulsar locations to be constant with respect to the Earth, i.e. $\bar{q}$ is not a function of time. In practice, the pulsar locations vary with respect to the Earth, but are constant with respect to the solar system barycentre (SSB). This barycentering correction is typically applied during the pulsar detection process. Our conclusions are generally unchanged by this choice - it is straightforward to transform $q$ if needed into a vector between the SSB and pulsar. \newline

\noindent Bringing this all together we can write $\dot{p}_t$ in a condensed form as,
\begin{equation}
	\dot{p}_t = A e^{i B t + C}
\end{equation}

>>>>

% Title of the paper, and the short title which is used in the headers.
% Keep the title short and informative.
	\title[Kalman PTA]{Kalman tracking and estimation of continuous gravitational waves with a pulsar timing array}

@@

% Select between one and six entries from the list of approved keywords.
% Don't make up new ones.
\begin{keywords}
gravitational waves -- methods: data analysis -- pulsars: general
\end{keywords}

@@

% The MNRAS class isn't designed to include a table of contents, but for this document one is useful.
% I therefore have to do some kludging to make it work without masses of blank space.
\begingroup
\let\clearpage\relax
%\tableofcontents
\endgroup
\newpage
\section{Introduction}\label{sec:intro}
The inspiral of supermassive black hole binaries \citep[SMBHBs;][]{Rajagopal1995,Jaffe_2003, Wyithe2003,Sesana2013,McWilliams_2014,Ravi2015MNRAS.447.2772R,Burke2019, Skyes2022} is predicted to emit nHz gravitational waves (GWs). Other GW sources in this low-frequency regime include cosmic strings \citep[e.g.][]{PTAstring} and cosmological phase transitions \citep[e.g.][]{PTAphase}. The detection of nHz GWs has necessitated the development of new observational methods, since it is practically impossible to engineer interferometric detectors with sufficiently long baselines. The foremost method is timing an ensemble of pulsars; a pulsar timing array \citep[PTA;][]{ Tiburzi2018, 2021hgwa.bookE...4V}. A nHz GW influences the trajectory and frequency of radio pulses, leaving a characteristic impression on the pulse times of arrival (TOAs) measured at the  Earth. By measuring TOAs from multiple pulsars simultaneously one can effectively construct a detector with a baseline on the scale of parsecs. Multiple PTA detectors have been built over the last few decades, including the North American Nanohertz Observatory for Gravitational Waves \citep[NANOGrav,][]{NANOgrav2023}, the Parkes Pulsar Timing array \citep[PPTA][]{Parkes2023}, and the European Pulsar Timing Array \citep[EPTA,][]{EPTA2023}. These individual efforts have joined in international collaboration, under the umbrella of the International Pulsar Timing Array \citep[IPTA,][]{2019MNRAS.490.4666P}, along with a number of newer PTAs such as the Indian Pulsar Timing Array Project \citep[InPTA,][]{ipta}, MeerTime \citep{meertime2,Meertime} and the Chinese PTA \citep{Hobbs_2019}. \newline

\textcolor{red}{TK: Content and references need to be updated in light of recent PTA results}. The incoherent superposition of multiple SMBHB sources leads to a stochastic GW background detectable at nHz frequencies \citep{Allen1997,Sesana10,Christensen2019,Renzini2022}. Previous efforts have mainly focused on detecting the stochastic background by measuring the cross-correlation between the pulsar timing residuals between pairs of pulsars as a function of the angular separation between the pulsars -- the Hellings-Downs curve \citep{Hellings}. However, no GW background has yet been detected \citep{Lentati2015,NanoGrav2018,2022MNRAS.510.4873A}. \newline

Individual SMBHBs that are sufficiently massive and nearby may be resolvable with PTAs, allowing the very earliest stages of their evolution and coalescence to be investigated \citep{Sesana2010,Yardley2010,Zhu10,Babak2012,2013CQGra..30v4004E,Zhupulsarterms}. 
Indeed, the stochastic GW background itself may be dominated by a few individual binary sources \citep{Ravi2012singlesource}. Individual SMBHBs are continuous wave sources: they generate persistent, quasi-monochromatic modulations of a known form in pulsar timing residuals. Consequently, they are detected more efficiently by either a frequentist matched filter e.g.\ the ${\cal F}$-statistic \citep{Lee2011MNRAS.414.3251L, Ellis2012ApJ,Zhu2014PPTA} or else Bayesian inference \citep{Ellis2016,Arzoumanian2020A} rather than by cross-correlating pulsar pairs. However, PTA observational efforts to detect individual sources have thus far been unsuccessful \citep{Jenet2004,Zhu2014PPTA,Babak2016,Arzoumanian2023}. \newline

Intrinsic pulsar timing noise -- i.e.  random, unmodelled, red-spectrum TOA fluctuations due to irregularities in the rotation of the star -- has been identified as a key factor limiting the sensitivity of PTAs to GW signals \citep{Shannon2010,Lasky2015,Caballero2016,Goncharov2021}. This timing noise has multiple theorized causes including free precession \citep{free_precession_kerr,stairs_freeprecession}, microglitches \citep{Alessandro1995,Melatos2008,Espinoza2021}, asteroid encounters \citep{Shannon_2013,Brook_2014}, glitch recovery \citep{Johnston10,Hobbs2010glitch}, fluctuations in internal and external stochastic torques \citep{Cordes1981, 2006MNRAS.370L..76U,Antonelli2023}, variations in the coupling between the stellar crust and core \citep{Jones1990MNRAS.246..364J,Meyers2021,Melatos2023}, magnetospheric state switching \citep{magneto1,Lyne2010L,Stairs2019MNRAS.485.3230S} and superfluid turbulence \citep{Greenstein1970,Peralta2006,Melatos2014}. In order to mitigate the impact of timing noise, PTAs are typically composed of millisecond pulsars (MSPs) which are relatively stable rotators. However, timing noise in MSPs may be a latent phenomenon that will increasingly assert itself as longer stretches of more sensitive data are analysed in the quest to detect nHz GWs \citep{Shannon2010}. In modern Bayesian PTA searches, the power spectral density of the intrinsic timing noise is modeled (usually as a broken or unbroken power law) and estimated, in an effort to distinguish it from the red noise induced by a stochastic GW background (whose spectrum is also red). In addition to the red timing noise there are also secondary, white noise sources that must be considered such as phase jitter noise and radiometer noise \citep{Cordes2010,Lam2019,Parthasarathy2021}. \newline

In this work we present an alternative and complementary approach to PTA data analysis for individual, quasi-monochromatic, SMBHB sources which self-consistently tracks the intrinsic timing noise in PTA pulsars and disentangles it from GW-induced TOA modulations. The new approach differs from existing approaches in one key respect: it infers the GW parameters conditional on the unique, time-ordered realization of the noisy TOAs observed, instead of fitting for the ensemble-averaged statistics of the TOA noise process, e.g., the amplitude and exponent of its power spectral density. Stated another way, existing approaches seek to detect a GW signal by marginalizing over the ensemble of possible noise realizations, whereas the new approach delivers the most likely set of GW parameters consistent with the actual, observed noise realization. The new and existing approach are therefore complementary, but the new approach holds out the promise of somewhat higher sensitivity --- a promise which we aim to test as the key goal of this paper. In particular, we formulate PTA analysis as a state-space problem and demonstrate how to optimally estimate the state-space evolution using a Kalman filter, a tried-and-tested tool \citep{Kalman1,Meyers2021,Melatos2023}. We combine the Kalman tracking of the pulsars intrinsic rotational state with a Bayesian nested sampler \citep{Skilling, Ashton2022} to estimate the GW parameters and calculate the marginal likelihood for model selection.\textcolor{red}{TK: a few extra sentences here on how we compare to existing approaches. Not quite sure yet how to do this.} \newline

\noindent This paper is organised as follows. In Section \ref{sec:model} we present the state-space model for the pulsar pulse frequency which is subject the influence of a GW. In Section \ref{sec:detect} we discretise the model, develop a Kalman filter to track the state evolution and introduce how to deploy the Kalman filter in conjunction with nested sampling to estimate the system parameters and the model evidence. \textcolor{red}{TK: this paper outline section will be completed once we have finalized the rest of the manuscript} In Section \ref{sec:testing} we test this new method on synthetic pulsar data. Throughout this work we adopt the natural units, with $c = G = \hbar = 1$, and a $(-,+,+,+)$ metric signature. \newline

@@

\section{State-Space Formulation}\label{sec:model}
We formulate the PTA analysis as a state-space problem, in which the intrinsic rotational state of each pulsar evolves according to a stochastic differential equation and is related to the observed pulse sequence via a measurement equation. In this work we take the intrinsic state variable to be the pulsar's spin frequency $f_{\rm p}^{(n)}(t)$, as measured in the local rest frame of the pulsar's centre of mass. A model for the evolution of $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_frequency}.  We take the measurement variable to be the radio pulse frequency measured by an observer at Earth, $f_{\rm m}^{(n)}(t)$.  The measurement equation relating $f_{\rm m}^{(n)}(t)$ to $f_{\rm p}^{(n)}(t)$ is presented in Section \ref{sec:psr_measured}. The superscript $1\leq n\leq N$ indexes the $n$-th pulsar in the array.
\subsection{Spin evolution} \label{sec:psr_frequency}

A predictive, first-principles theory of timing noise does not exist at present; there are several plausible physical mechanisms, referenced in Section \ref{sec:intro}. We therefore rely on an idealized phenomenological model to capture the main qualitative features of a typical PTA pulsar's observed spin evolution, i.e.\ random, small-amplitude excursions around a smooth, secular trend. In the model, $f_{\rm p}^{(n)}(t)$ evolves according to the sum of a deterministic torque and a stochastic torque. The deterministic torque is attributed to magnetic dipole braking, with braking index $n_{\rm em}=3$. Most PTAs involve millisecond pulsars, for which the quadratic correction due to $n_{\rm em}$ in $f_{\rm p}^{(n)}(t)$ is negligible over the observation time $T_{\rm obs} \sim 10 \, {\rm yr}$, and the deterministic evolution can be approximated accurately by 
\begin{equation}
 f_{\rm em}^{(n)}(t) = f_{\rm em}^{(n)}(t_1) + \dot{f}_{\rm em}^{(n)}(t_1)t \ , \label{eq:spinevol}
\end{equation} where an overdot denotes a derivative with respect to $t$ and $t_1$ labels the time of the first TOA. The stochastic torque is assumed to be a zero-mean white noise process. Specifically, the frequency evolves according to an Ornstein-Uhlenbeck process, described by a Langevin equation with a time-dependent drift term \citep{Vargas}
\begin{equation}
	\frac{df_{\rm p}^{(n)}}{dt} = -\gamma^{(n)}	 [f_{\rm p}^{(n)} - f_{\rm em}^{(n)} (t)] + \dot{f}_{\rm em}^{(n)} +\xi^{(n)}(t) \ . 
	\label{eq:frequency_evolution}
\end{equation}
In Eq \ref{eq:frequency_evolution} $f_{\rm EM}$ is the solution of the electromagnetic spindown equation, $\dot{f}_{\rm EM}$ is the spin derivative, $\gamma^{(n)}$ a proportionality constant whose reciprocal specifies the mean-reversion timescale, and $\xi^{(n)}(t)$ satisfies:
\begin{equation}
	\langle \xi^{(n)}(t) \rangle = 0 \ ,
\end{equation}
\begin{equation}
	\langle \xi^{(n)}(t) \xi^{(n)}(t') \rangle = [\sigma^{(n)}]^2 \delta(t - t') \ .
	\label{eq:xieqn}
\end{equation}
$[\sigma^(n)]^2$ is the variance of $\xi^{(n)}$ and parametrizes the amplitude of the noise, and combined with the mean reversion it gives a characteristic root mean square fluctuations $\approx [\sigma^{(n)}]^2 / \gamma^{(n)}$ in $f_{\rm p}^{(n)}(t)$ \citep{gardiner2009stochastic}. It is important to note that the white noise fluctuations in $\xi(t)$ translate into red noise fluctuations in the rotational phase $\phi(t) = \int_{t_1}^t dt' \, f_{\rm p}(t')$ after being filtered by the terms involving $d/dt$ and $\gamma$ in Eq. \ref{eq:spinevol}, consistent with the observed power spectral density of typical millisecond pulsars in the nHz band relevant to PTA experiments. \newline

Equations \ref{eq:spinevol}--\ref{eq:xieqn} represent a phenomenological model, which aims to reproduce qualitatively the typical timing behaviour observed in PTAs, viz.\ a mean-reverting random walk about a secular spin-down trend [ref latest results paper from each of the three main PTAs]. Equations \ref{eq:spinevol}--\ref{eq:xieqn} are not derived from first principles by applying a microphysical theory. As a first pass, they also exclude certain phenomenological elements, which are likely to be present in reality, e.g.\ the classic, two-component, crust-superfluid structure inferred from post-glitch recoveries \citep{Baym1969,vanEysden,Alpar2017MNRAS.471.4827G}. An approach akin to Equations \ref{eq:spinevol}--\ref{eq:xieqn} has been followed successfully in other timing analyses in the context of anomalous braking indices \citep{Vargas} and hidden Markov model glitch searches \citep{Melatos2020ApJ...896...78M,Lower2021MNRAS.508.3251L,Dunn2022,Dunn2023MNRAS.522.5469D}. However, Equations \ref{eq:spinevol}--\ref{eq:xieqn}  involve significant idealizations, which must be recognized at the outset \citep{Meyers2021,Myers2021MNRAS.502.3113M,Vargas}. First, the white noise driver $\xi(t)$ in Equation \ref{eq:frequency_evolution} is not differentiable, which makes the formal interpretation of $d^2 f_{\rm p} / dt^2$ ambiguous, even though $d^2 f_{\rm p} / dt^2$ is not used in the PTA analysis proposed in this paper. Second, the white spectrum assumed for $\xi(t)$ may or may not be suitable for millisecond pulsars in PTAs. It is challenging observationally to infer the spectrum of $\xi(t)$ from the observed spectrum of the phase residuals, because the inference is conditional on the (unknown) dynamical model governing $df_{\rm p}/dt$. For small-amplitude fluctuations sampled relatively often, as in millisecond pulsars in PTAs, it is likely that $\xi(t)$ is white to a good approximation over the inter-TOA intervals and generates red phase residuals as observed, but caution is warranted nevertheless. Third, the Brownian increment $dB(t)=\xi(t)dt$ does not include non-Gaussian excursions such as L\'{e}vy flights \citep{Sornette2004} which have not been ruled out by pulsar timing experiments to date. The above three idealizations are supplemented by other, physical approximations noted above, e.g.\ neglecting $n_{\rm em}$ in Equation \ref{eq:spinevol} and differential rotation between the crust and superfluid in Equation \ref{eq:frequency_evolution}.

\subsection{Modulation of pulse TOAs by a GW} \label{sec:psr_measured}
In the presence of a GW, the pulse frequency measured by an observer in the local rest frame of the neutron star's center of mass is different from that measured by an observer on Earth. Indeed, the pulse TOAs are modulated harmonically at the GW frequency. We derive the nonlinear measurement equation relating $f_{\rm m}(t)$ to $f_{\rm p}(t)$ in this section. The measurement equation is a key input into the Kalman filter in Section \ref{sec:kalman_filter}
\subsubsection{Plane GW perturbation}
We consider a gravitational plane wave from a single,distant source, which perturbs a background Minkowski metric $\eta_{\mu \nu}$ as
\begin{equation}
	g_{\mu \nu} = \eta_{\mu \nu} + H_{\mu \nu} \exp{[i(\Omega(\boldsymbol{n} \cdot \boldsymbol{x} - t) + \Phi_0)]} \ ,
\end{equation}
with spatial coordinates $\boldsymbol{x}$ and global coordinate time $t$. The GW has a constant angular frequency $\Omega$, propagates in the $\boldsymbol{n}$-direction (where $\boldsymbol{n}$ is a unit vector), has amplitude tensor $H_{\mu \nu}$, and has a phase offset  $\Phi_0$. Throughout this paper we will work with pulsar TOAs which have been defined relative to the Solar System barycentre (SSB). We are free to choose our coordinate system such that $\Phi_0$ is the GW phase at $t=0$ at the SSB. In this paper $\Omega$ has no time dependence; the source is approximated as monochromatic. Studies of SMBHB inspirals in the PTA context show that the gravitational wave frequency $f_{\rm gw}$ ($=\Omega / 2 \pi $) evolves over decadal timescales as \citep[e.g.][]{Zhu10},
\begin{equation}
	\Delta f_{\rm gw} \simeq 3.94 \mathrm{nHz}\left(\frac{M_c}{10^9 M_{\odot}}\right)^{5 / 3}\left(\frac{f_{\rm gw}(t=0)}{10^{-7} \mathrm{~Hz}}\right)^{11 / 3}\left(\frac{T_{\mathrm{obs}}}{10 \mathrm{yr}}\right)
	\label{eq:f_evolution}
\end{equation}
where $M_c$ is the chirp mass of the SMBHB, $f_{gw}(t=0)$ is the GW frequency at the time of the first observation, and $T_{\rm obs}$ the length of the data timespan, which for PTAs is $\sim 10$ years. A source can be considered as monochromatic if $\Delta f_{\rm gw}$ is less than the PTA frequency resolution of $1/T_{\rm obs}$. From Eq. \ref{eq:f_evolution} we can see that only those binaries which are very massive or at very high frequency experience significant frequency evolution over typical PTA timespans. The majority of SMBHBs detectable with PTAs are expected to satisfy $\Delta f_{\rm gw} < 1/T_{\rm obs}$; for a PTA with pulsars at 1.5 kpc, 78 \% of simulated SMBHBs satisfy this condition for the current IPTA, whilst for the second phase of the Square Kilometer Array (SKA2) this fraction drops to 52 \% \citep[Fig 7 of ][]{Rosado10.1093/mnras/stv1098}. We are therefore justified in treating the GW source as monochromatic \citep{Sesana10,Sesana2010,Ellis2012ApJ}. \newline

The amplitude tensor $H_{\mu \nu}$ has zero temporal components ($H_{0 \mu} = H_{\mu 0} = 0$). The spatial part is
\begin{align}
	H_{ij} = h_+ e_{ij}^+(\boldsymbol{n}) + h_{\times} e_{ij}^{\times}(\boldsymbol{n}) \ ,
\end{align}
where $h_{+}$ and $h_{\times}$ are the respective polarisation amplitudes. The plus and cross polarisation tensors $e_{ij}^{+}$ and $e_{ij}^{\times}$ are uniquely defined by the principal axes of the wave, viz.\ the unit 3-vectors $\boldsymbol{k}$ and $\boldsymbol{l}$, according to
\begin{align}
	e_{i j}^{+}(\boldsymbol{n}) =k_i k_j-l_i l_j \ , \\
		e_{i j}^{\times}(\boldsymbol{n}) =k_i l_j+l_i k_j \ .
\end{align}
The principal axes are in turn specified by the location of the GW source on the sky (via colatitude $\theta$ and azimuth $\phi$) and the polarisation angle $\psi$ according to
\begin{align}
	\boldsymbol{k}  = &(\sin \phi \cos \psi-\sin \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
	& -(\cos \phi \cos \psi+\sin \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}} \nonumber \\
	& +(\sin \psi \sin \theta) \boldsymbol{\hat{z}} \\
	\boldsymbol{l} = &(-\sin \phi \sin \psi-\cos \psi \cos \phi \cos \theta) \boldsymbol{\hat{x}} \nonumber \\
	& +(\cos \phi \sin \psi-\cos \psi \sin \phi \cos \theta) \boldsymbol{\hat{y}}\nonumber  \\
	& +(\cos \psi \sin \theta) \boldsymbol{\hat{z}}
\end{align}
where e.g. $\boldsymbol{\hat{x}}$ is a unit vector in the direction of the $x$-axis. The direction of GW propagation is related to the principal axes by
\begin{equation}
	\boldsymbol{n} = \boldsymbol{k} \times \boldsymbol{l} \ . 
\end{equation}

\subsubsection{Measurement equation}
\textcolor{red}{TK: Does this subsection needs a better name?} \newline 
\noindent \textcolor{red}{TK: this subsection has been overhauled per comments from AM. The structure is now: \newline 1. Radio pulses described by general geometric object $p$ \newline 2. Shift in $p$ due to a GW (no derivation, just definition of terms) \newline 3.Relate shift in $p$ to a shift in $f$ by clarifying that $u=(1,0,0,0)$ for PSR and observer. \newline 4. Give final measurement equation} \newline
